# **Course Project I**

===================

## Loading the data

```{r, echo = TRUE}
data <- read.csv("C:/Users/ddelp/OneDrive/Escritorio/activity.csv")
```


## Exploring the data

```{r, echo = TRUE}
str(data)
summary(data)
```


## Total number of steps taken each day

We remove missing values first and then get the number of steps taken each day, mean and median

```{r, echo = TRUE}
library(dplyr)
data1 <- data[complete.cases(data), ]
results1 <- group_by(data1, date) %>% 
  summarise(total = sum(steps), mean = mean(steps), median = median(steps))
print (results1, n = 53)
```


## Mean and median total number of steps taken per day

```{r, echo = TRUE}
mean(results1$total)
median(results1$total)
```


## Histogram of the total number of steps taken each day

```{r histogram1, echo = TRUE}
hist(results1$total, col = "skyblue", breaks = 20, main = "Total number of steps taken each day",
     xlab = "Steps")
```


## Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo = TRUE}
av_interval <- group_by(data1, interval) %>%
  summarise(mean_steps = mean(steps))
av_interval <- as.data.frame(av_interval)
```

First we group the data frame by interval and take the mean of "steps" variable.

```{r, echo=TRUE}
library(ggplot2)
plot1 <- ggplot(av_interval, aes(x = interval, y = mean_steps))
plot1 + geom_line(colour = "salmon2", size = 1) + labs (x = "5-min interval", 
                                                        y = "Average number of steps") +
  ggtitle("Time series plot: average number of steps across days")

```


## The 5-minute interval that, on average, contains the maximum number of steps

We use the data frame grouped by interval, then reorder our data so that the maximum value is first so we can select it.

```{r, echo = TRUE}
max_interval <- av_interval %>%
  arrange(desc(mean_steps)) %>%
  top_n(1) %>%
  select(interval)

print(max_interval)
```


## Code to describe and show a strategy for imputing missing data


### Total number of missing values in the data set
```{r, echo = TRUE}
summary(data)
sum(is.na(data$date))
sum(is.na(data$interval))
sum(is.na(data$steps))
```
There are 2304 NA values and only in variable "steps".

### Devise a strategy for filling in all of the missing values in the dataset. 
We are going to replace missing values of variable "steps" with the calculated mean by interval, in the data frame without missing values.
We will call this variable "steps_hat"

```{r, echo = TRUE}
imput_steps = group_by(data1, interval) %>%
  summarise(steps_hat = mean(steps))
imput_steps = as.data.frame(imput_steps)
```


### Create a new dataset that is equal to the original dataset but with the missing data filled in.
We merge our original data frame with the "steps_hat" variable and for those cases where there is a missing value, the variable "steps" will be substituted by the value of "steps_hat" (our imputed mean).

```{r, echo = TRUE}
data2 = left_join(data, imput_steps, by = "interval")
data2$steps_final = ifelse(is.na(data2$steps) == TRUE,
                           round(data2$steps_hat), 
                           data2$steps)
head(data2)
sum(is.na(data2$steps_final))
```
We check that there are no missing values.


### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r histogram2, echo=TRUE}
results2 <- group_by(data2, date) %>%
  summarise(total2 = sum(steps_final))
hist(results2$total2, col = "cyan", breaks = 20, main = "Total number of steps taken each day",
     xlab = "Steps")
mean(results2$total2)
median(results2$total2)
```
The mean is the same as the mean from the first part of the assignment, and the median did not change much. Given that a lot of values are replaced with the mean, it makes sense that mean and median are similar. 

## Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

### First create the factor variable weekend.

```{r, echo=TRUE}
data2$date = as.Date(data2$date)
data2$weekend = weekdays(data2$date)
table(data2$weekend)

data2$weekend = factor (ifelse(data2$weekend %in% c("sábado","domingo"), 1, 0),
                        labels=c('Weekday', 'Weekend'))
table(data2$weekend)
```

### Panel plot

```{r panelplot, echo=TRUE}
results3 <- data2 %>%
  group_by(interval, weekend) %>%
  summarise(mean3 = mean(steps_final))
plot2 <- ggplot(results3, aes(interval, mean3))
plot2 + geom_line(colour = "salmon2", size = 1) + facet_grid(weekend ~ .) + labs(x = "5-min interval",
                                                                                 y = "Average number of steps") + ggtitle("Time series plot: Average number of steps taken, by weekend or weekday")
```


